TriColorings of Knots

### 2018-04-06 paper/paper.tex
Finished first draft of paper

%% Revised 11/12/05.
%% Notice that anything written on a line that begins 
%% with  a % sign is ignored by LaTeX.
\documentclass[12pt]{amsart}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

@@

\newcommand{\bb}[1]{\mathbb{#1}}

@@

\begin{abstract}
  Empty
\end{abstract}

\maketitle

@@

\textbf{Citation Usage EX}~\cite[Problem 2.4.6]{jpsk}

@@

@@

\begin{example}
  The trefoil knot has three crossings and three arcs. Label it as
  shown in figure~\ref{fig:AnnTref}. Then the Wirtinger presentation
  of the knot group of the trefoil is
  \[
    \langle x_1,x_2,x_3|x_1x_2x_1^{-1}=x_3,x_3x_1x_3^{-1}=x_2,x_2x_3x_2^{-1}=x_1\rangle
  \]
  However the presentation of this group can be simplified to the
  form
  \[
    \langle x_1,x_1| x_1x_2x_1=x_2x_1x_2\rangle
  \]
\end{example}

\begin{figure}
  \caption{Annotated Trefoil}
  \label{fig:AnnTref}
\end{figure}

>>>>

\documentclass[12pt]{amsart}

\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}

\graphicspath{{images/}}

@@

\newcommand{\bb}[1]{\mathbb{#1}}
\newcommand{\ds}{.3}

@@

\begin{abstract}
  Tricoloring is a knot invariant that differentiates knots
  based on whether or knot their diagram can be colored in a
  certain way with 3 colors. In this paper we introduce knots
  and define Tricolorability in terms of both knot diagrams
  and the knot group. We then briefly mention extensions
  of the notions of tricolorability to arbitrary $n$-coloring.
\end{abstract}

\maketitle
\textbf{TODO}:
\begin{enumerate}
\item Remove \textbf{TODO}
\item Iron out figure/image placement before final edition. Too early
  now. Wait till revision is done.
\end{enumerate}

@@

One common pattern we see in mathematics is a concern with how
certain objects fit inside other objects. Knot theory studies this
sort of problem. Given a circle $S^1$ what different ways can we embed
it into $\bb{R}^3$ or $\bb{R}^3$  with an additional point at infinity ($S^3$)?
A typical drawing of a knot can be seen in figure~\ref{fig:figure-eight}.

\begin{figure}
  \includegraphics[scale=\ds]{figure-eight}
  \caption{Figure-eight Knot}
  \label{fig:figure-eight}
\end{figure}

Naturally once we begin to consider knots we are also led to consider
how we might tell them apart or rather when they should be the same
at all. Intuitively the knot in figure~\ref{fig:figure-eight} should
be considered the same as in~\ref{fig:figure-eight-t} but should be
different than that of trefoil seen in figure~\ref{fig:trefoil}. In
the next section we will see the definition of a knot, how to define their
equivalence in a sensible way, followed by a method to differentiate them
called tricolorability. Afterwards we will look at an alternative definition
of tricolorability. Then we will introduce notions of how one may consider
coloring knots with more than 3-colors.

\begin{figure}
  \includegraphics[scale=\ds]{figure-eight-t}
  \caption{Figure-eight with a twist}
  \label{fig:figure-eight-t}
\end{figure}

@@

In this paper we will follow the conventions of Charles
Livingston~\cite{Livingston}. To begin we give the definition of a knot.

\begin{definition}[Knot]
  A Knot is a simple closed curve in $\bb{R}^3$(or $S^3$).
\end{definition}

Now there are two issues that we need to consider before moving forward.
The first is our notion of equivalence. Unfortunately a homeomorphism from
one knot to another is too weak a notion of equivalence. Instead we will
use an equivalence called ambient isotopy.

\begin{definition}[Ambient Isotopy]
  Two knots $K,J$ are ambient isotopic if there is a homotopy on the ambient
  space $H:S^3\times [0,1]\rightarrow S^3$ if $H(x,0)$ is the identity and the
  image $H(K,1)=J$.
\end{definition}

Second is that we will be considering a subset of knots called tame. A knot
is tame it is ambient isotopic to a piecewise linear knot. An example of
a knot that is not tame is called wild and an example drawing of such can be
seen in figure~\ref{fig:wild}. Wild knots have some more ``pathological''
properties and in order to keep things simple and consistent.

\begin{figure}
  \includegraphics[scale=\ds]{wild}
  \caption{Wild Knot}
  \label{fig:wild}
\end{figure}

When we give the initial definition of  tricolorability we will not work
directly with the curves themselves. Instead we will deal with diagrams
of the knots. All drawings of knots in this paper are examples of diagrams.

\begin{definition}[Diagram]
  A diagram $D$ for a knot $K$ is a projection of $K$ down to a 2-plane
  such that no three points are sent to the same place nor do any two
  points that are kinks in $K$.
\end{definition}

There is a loss of information that occurs when projecting down to a 2-plane
and we recover this by taking any intersection and denoting which strand
goes above and which strand goes below. The segments that make up the
diagram are called arcs.

There is one other unfortunate occurrence
from moving to diagrams. It is that there are infinitely many diagrams
for a given knot. Moreover actually telling whether to diagrams denote
the same knot is extremely difficult although it is decidable and moreover
belongs to NP~\cite{Poonen}.

It turns out that there are three ``moves'' along with planar isotopy
that we can apply to diagrams that will remedy this.

\begin{definition}[Reidmeister Moves]
  The Reidmeister moves are the moves labeled in figure~\ref{fig:reidmeister}.
  We will refer to them as type I, type II, or type III moves respectively.

  \begin{figure}
    \includegraphics[scale=\ds]{reidmeister}
    \caption{Reidmeister moves}
    \label{fig:reidmeister}
  \end{figure}
\end{definition}

We will use the following theorem about Reidmeister moves without proof.

\begin{theorem}
  Two knots $K$ and $J$ are ambient isotopic if and only if their diagrams
  are related by a finite sequence of Reidmeister moves and planar isotopy.
\end{theorem}

This is where the importance of Reidmeister moves and diagrams becomes apparent
in defining invariants of knots. If we define an invariant on diagrams and show
that it does not change under planar isotopy or the Reidmeister moves then
it will also act as an invariant on knots. This is precisely the method we
will use with our initial definition of tricolorability.

Now there are two knots that we will return to over an over again in this paper
as they are the simplest examples of knots for which tricolorability
differentiates them. The simplest knot is the unknot~\ref{fig:unknot} which
is unique in that it has a diagram with no crossings whatsoever. The
other is the trefoil~\ref{fig:trefoil}.

\begin{figure}
  \includegraphics[scale=\ds]{unknot}
  \caption{Unknot}
  \label{fig:unknot}
\end{figure}

\begin{figure}
  \includegraphics[scale=\ds]{trefoil}
  \caption{Trefoil Knot}
  \label{fig:trefoil}
\end{figure}

With that in mind let us introduce tricolorability.

\begin{definition}
  Let $K$ be a knot with diagram $D$. Then a tricoloring (3-coloring)
  of $K$ is an assignment to each arc one of three colors
  such that at each crossing either all colors or
  only single color color appear. A tricoloring is called nontrivial
  if at least 2 colors are used.
\end{definition}

We will use the colors red, green, and blue as our colors.
Now we must show that tricoloring is in fact an invariant.

\begin{theorem}
  The existence of a non-trivial 3-coloring for some diagram is a
  knot invariant.
\end{theorem}

\begin{proof}
  Let $K$ be a knot with diagram $D$. We must show that given a coloring
  for $D$ that we can color any diagram of $K$. First note that planar isotopy
  does not affect the crossings for $D$ and as such does not change the
  3-colorability.

  Next we will have a series of pictures that demonstrate the possibilities
  for Type I, Type II, and Type III moves respectively.

  For type I we have:
  \begin{figure}
    \includegraphics[scale=\ds]{t1-c}
    \caption{Type I}
    \label{fig:t1-c}
  \end{figure}

  For type II there are two cases either both strands have the same color
  or they are different.
  \begin{figure}
    \includegraphics[scale=\ds]{t2-c}
    \caption{Type II}
    \label{fig:t2-c}
  \end{figure}

  Finally for type III moves there are several cases based on
  the colors of the strands entering. We will demonstrate
  three such cases and the rest are similar.
  \begin{figure}
    \includegraphics[scale=\ds]{t3-c}
    \caption{Type III}
    \label{fig:t3-c}
  \end{figure}

  Thus if a diagram $D$ for a knot $K$ is has a 3-coloring then
  any diagram for $K$ has a 3-coloring. Therefore 3-colorability
  is a knot invariant.
\end{proof}

An immediate consequence of this is that the unknot and the trefoil
knot are in fact not identical. We can see this because the unknot
is not tricolorable as it only has a single arc. However the
trefoil can be colored as in figure~\ref{fig:trefoil-c}.

\begin{figure}
  \includegraphics[scale=\ds]{trefoil-c}
  \caption{Colored trefoil}
  \label{fig:trefoil-c}
\end{figure}

@@

\begin{example}
  The trefoil knot has three crossings and three arcs. Label it as
  shown in figure~\ref{fig:annotated-trefoil}. Then the Wirtinger presentation
  of the knot group of the trefoil is
  \[
    \langle x_1,x_2,x_3|x_1x_2x_1^{-1}=x_3,x_3x_1x_3^{-1}=x_2,x_2x_3x_2^{-1}=x_1\rangle
  \]
  However the presentation of this group can be simplified to the
  form
  \[
    \langle x_1,x_1| x_1x_2x_1=x_2x_1x_2\rangle
  \]
\end{example}

\begin{figure}
  \includegraphics[scale=\ds]{annotated-trefoil}
  \caption{Annotated Trefoil}
  \label{fig:annotated-trefoil}
\end{figure}
